### 2013-06-11 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding some example figures for domain decomposed monte carlo

The neutron transport problem is complicated. Solutions cover a large
phase space and the problems of interest are often geometrically
complex, very large, or both, requiring tremendous computational
resources to generate an adequate solution. Modern deterministic
methods for large scale problems are commonly variants on the discrete
ordinates ($S_N$) method \citep{evans_denovo:_2010}. For fission
reactor neutronics simulations, the $S_N$ method requires potentially
trillions of unknown angular flux moments to be computed to achieve
good accuracy for the responses of interest
\citep{slaybaugh_acceleration_2011}. Other forms of the transport
problem, including the $P_N$ method, take on a simpler form than the
more common $S_N$ methods but lack in accuracy when compared while
still requiring considerable computational resources for solutions in
multiple dimensions.

In the 1960's, Gelbard developed an ad-hoc multidimensional extension
of the simple single dimension planar $P_N$ equations that created a
system of coupled, diffusion-like equations known as the simplified
$P_N$ ($SP_N$) equations. Up until around the 1990's, the $SP_N$
method was either widely unknown, widely unused, or combination of
both even though numerical studies showed promising results with
better solutions than diffusion theory and a significant reduction in
computational time over more accurate methods such as discrete
ordinates. Why did this happen? A significant problem, pointed out by
Brantley and Larsen \citep{brantley_simplified_2000}, was that little
rigor had been applied to the formulation of the $SP_N$ equations
since their derivation through primarily heuristic arguments. Instead,
studies at that time focused on simply comparing the results of the
method to other contemporary transport solution strategies. In
addition, many problems of interest from the literature at the time
were either solved using nodal-type methods for reactor-sized problems
or $S_N$-type methods for benchmark problems with intricate material
configurations and potentially large flux gradients over small spatial
domains.

So why reconsider the $SP_N$ equations? Starting in the 1990's and
primarily due to Larsen and his colleagues, the $SP_N$ equations have
been given a more rigorous treatment with both variational and
asymptotic derivations performed as a means of verification. In
addition, these equations have been more rigorously studied as
solution methods to MOX fuel problems and have been shown to provide
accurate solutions. With this mathematical literature to provide a
solid numerical footing for the method, we look at its application to
today's challenge problems in neutron transport for fission reactor
analysis. The reduction in numerical complexity of current
deterministic solution methods using the $S_N$ approximation could
mean significant savings in both compute time and memory required. In
addition, the characteristics of the solution to the transport problem
for a steady state reactor core permit diffusion theory to be used; a
staple of the nuclear industry since its inception. Therefore, if
diffusion theory is applicable, then finer grained solutions that
capture more of the physics contained in the transport equation should
be possible with the $SP_N$ method. In doing so, we also expect from
the literature to obtain computed responses on the order of accuracy
we would expect from an appropriately discretized $S_N$ method at a
fraction of the cost.

To further motivate moving in this direction, recent developments in
the Exnihilo neutronics package at Oak Ridge National Laboratory have
permitted generation of the $SP_N$ system of equations for detailed
neutronics models \citep{evans_simpli_2013}. By fully forming these
equations and formulating them as a linear algebra problem instead of
using the explicit iterative methods of the past, we now have access
to all of the modern advancements in computational linear algebra
including Krylov solvers for asymmetric systems and preconditioning
methods such as algebraic multigrid. This leads us to then explore the
applicability of our work in discrete Monte Carlo methods for linear
systems as a possible solution method for the $SP_N$ equations. If
formulated correctly, we hypothesize that significant improvement in
usage of computational resources may be observed compared to modern
solution techniques such as those suggested due to the form of the
matrices generated by the discretization. In addition, solving the
$SP_N$ equations in this way also breaks away from the $S_N$ forms of
parallelism where spatial parallelism is achieved by an efficient
parallel sweep, angular efficiency achieved by pipe-lining, and energy
parallelism achieved by decoupling the groups. With the $SP_N$
equations as a full matrix system, we now can parallelize the problem
as prescribed by the linear solver, which may be significantly more
scalable than current $S_N$ transport practices.

In this chapter, we derive the $SP_N$ equations, closely following the
work of Evans\footnote{The $SP_N$ derivations in this chapter are
  heavily based on those presented by Evans in
  \citep{evans_simpli_2013}}, in order to gain full understanding of
the underlying system and its potential behavior in a Monte Carlo
context. From the $P_N$ equations, we apply a set of approximations to
yield the $SP_N$ equations for fixed source and criticality
problems. Using the fully-formed linear operator for the transport
problem, we explore solutions to the $SP_N$ equations using Monte
Carlo Synthetic Acceleration using a light water reactor fuel assembly
criticality calculation as the driving problem. In order to
effectively solve this problem, a suite of preconditioners is
developed and studied within the context of MCSA. Using these
preconditioners, MCSA solutions for the fuel assembly problem are
verified by comparing against production Krylov solvers and finally
performance is analyzed through comparison with those same Krylov
solvers.

@@

    u_4 = 7\phi_6 \:,

@@

or equivalently:
    \phi_0 = u_1 - \frac{2}{3}u_2 + \frac{8}{15}u_3 -
    \frac{16}{35}u_4 \\
    \phi_2 = \frac{1}{3}u_2 - \frac{4}{15}u_3 + \frac{8}{35}u_4\\ 
    \phi_4 = \frac{1}{5}u_3 - \frac{6}{35}u_4\\
    \phi_6 = \frac{1}{7}u_4\:.
  \label{eq:spn7_subs_inverse}

@@

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue from
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
source derived in \S~\ref{sec:pn_eigenvalue}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with the fission matrix $\mathbf{F}$ given by
Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
to yield the set of pseudo-moments $\mathbb{U}_n$ where now the
application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices.

It should be noted here that with respect to a general MCSA scheme,
the introduction of the fission in the system will only affect the
source vector in the linear system. The linear operator, and therefore
overall MCSA performance will be dictated by streaming and scattering
as defined on the left-hand side.

@@

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
bottom is given in Figure~\ref{fig:problem3_end}. On the top and
bottom of the assembly, vacuum conditions are used as well as on the
top and right boundaries in
Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homgenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of
MCSA. To investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary criteria needed
to satisfy MCSA convergence has also been met. The number of
iterations required for the Richardson iteration to converge will give
an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi
preconditioned\footnote{For 1-group $SP_1$ problems the block size is
  1, giving a point Jacobi preconditioner.}  calculations used a
one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use a simpler homogeneous 2-dimensional
one-speed neutron diffusion system to isolate this behavior. In this
system, we can vary the cross sections while maintaining a fixed grid
in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

\subsubsection{Model Diffusion Problem}
\label{subsec:model_problem}
For our breakdown study, we choose the one-speed, two-dimensional
neutron diffusion equation as a model problem
\citep{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

We choose a finite difference scheme on a square Cartesian grid to
discretize the problem. For the Laplacian, we choose the 9-point
stencil shown in Figure~\ref{fig:stencil} over a grid of size $h$
\citep{leveque_finite_2007}:
\begin{multline}
  \nabla^2_9\phi = \frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j}
    + 4 \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ +
    \phi_{i-1,j+1} + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20
    \phi_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.25}{\input{chapters/spn_equations/stencil.pdftex_t}}
  \end{center}
  \caption{\textbf{Nine-point Laplacian stencil.}}
  \label{fig:stencil}
\end{figure}
We then have the following linear system to solve:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{i-1,j} + 4 \phi_{i+1,j} + 4
    \phi_{i,j-1} + 4 \phi_{i,j+1} + \phi_{i-1,j-1}\\ + \phi_{i-1,j+1}
    + \phi_{i+1,j-1} + \phi_{i+1,j+1} - 20 \phi_{i,j}] + \Sigma_a
  \phi_{i,j} = s_{i,j}\:,
  \label{eq:fd_system}
\end{multline}
and in operator form:
\begin{equation}
  \ve{D}\boldsymbol{\phi}=\ve{s}\:,
  \label{eq:operator_system}
\end{equation}
where $\ve{D}$ is the diffusion operator, $\ve{s}$ is the source in
vector form and $\boldsymbol{\phi}$ is the vector of unknown fluxes.

To close the system, a set of boundary conditions is required. In the
case of a non-reentrant current condition applied to all global
boundaries of the domain, we choose the formulation of Duderstadt by
assuming the flux is zero at some ghost point beyond the
grid. Consider for example the equations on the $i=0$ boundary of the
domain:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{-1,j} + 4 \phi_{1,j} + 4 \phi_{0,j-1} +
    4 \phi_{0,j+1} + \phi_{-1,j-1}\\ + \phi_{-1,j+1} + \phi_{1,j-1} +
    \phi_{1,j+1} - 20 \phi_{0,j}] + \Sigma_a \phi_{0,j} = s_{0,j}\:.
  \label{eq:x_min_bnd}
\end{multline}
Here we note some terms where $i=-1$ and therefore are representative
of grid points beyond the boundary of the domain. We set the flux at
these points to be zero, giving a valid set of equations for the $i=0$
boundary:
\begin{multline}
  -\frac{1}{6h^2}[4 \phi_{1,j} + 4 \phi_{0,j-1} + 4 \phi_{0,j+1}
    \\ + \phi_{-1,j+1} + \phi_{1,j-1} + \phi_{1,j+1} - 20 \phi_{0,j}]
  + \Sigma_a \phi_{0,j} = s_{0,j}\:.
  \label{eq:x_min_bnd_2}
\end{multline}
We repeat this procedure for the other boundaries of the domain. For
reflecting boundary conditions, the net current across a boundary is
zero.

\subsubsection{MCSA Breakdown Analysis}
\label{subsubsec:breakdown_analysis}
For each solver and estimator combination, the spectral radius of the
iteration matrix generated by the diffusion problem was varied by
changing the absorption cross section from 0.25 to 100 while fixing
the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
scattering cross section of unity. For each parameter variation, a
minimum of one stochastic history per degree of freedom (DOF) in the
problem was used to compute the Monte Carlo correction. If the solver
could not converge in less than 100 iterations, the number of
histories was increased by increments of 5,000 until convergence was
achieved in less than 100 iterations. The number of iterations
required to converge MCSA and the time to converge was recorded as a
means to capture the breakdown.

@@

In addition to the significantly larger number of histories required
for to achieve convergence for ill-conditioned problems another
penalty is paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. We
therefore require a more expansive set of preconditioning techniques
to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
which MCSA is more applicable and in which performance is improved.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

@@

MCSA preconditioned with SPAINV-ILUT was also used to solve the same
fuel assembly problem as with the other preconditioners for the same
1-group $SP_1$ discretization. To study convergence sensitivity to
SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
were parametrically varied with the number of iterations to converge a
single eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. SPAINV parameters for the data presented here were fixed as
it was found that the preconditioning quality was largely insenstive
to their variation. In addition, the fact that SPAINV is being used to
preconditioned a problem already preconditioned with ILUT means that
the input matrix sparsity pattern is already somewhat dense and large
SPAINV levels would result in an even denser system. Therefore, a
smaller SPAINV sparsity pattern level will be used for this
preconditioning. For each calculation, the number of iterations
required to converge reported was for a single eigenvalue iteration.
Again, \sn{3}{4} histories are used at each MCSA iteration to compute
the Monte Carlo correction using the adjoint collision estimator. All
SPAINV-ILUT calculations reported here were performed on a single CPU.

>>>>

In this chapter, we briefly derive the simplified $P_N$ ($SP_N$)
equations, closely following the work of Evans
\citep{evans_simpli_2013}\footnote{The $SP_N$ derivations in this
  chapter and the associated appendices are heavily based on those
  presented by Evans in \citep{evans_simpli_2013}}, in order to gain
full understanding of the underlying system and its potential behavior
in a Monte Carlo context. From the $P_N$ equations, we apply a set of
approximations to yield the $SP_N$ equations for fixed source and
criticality problems. Using the fully-formed linear operator for the
transport problem, we explore solutions to the $SP_N$ equations with
Monte Carlo Synthetic Acceleration using a light water reactor fuel
assembly criticality calculation as the driving problem. Several
difficulties arise when applying MCSA to these problems that were not
observed when investigating the simpler transport systems.

For Jacobi-based preconditioners, convergence with MCSA is difficult
if not impossible for ill-conditioned systems with this behavior
demonstrated using a simple neutron diffusion problem. In order to
effectively solve the $SP_N$ equations for the fuel assembly problem,
a suite of preconditioners along with a relaxation scheme is developed
and studied within the context of MCSA. Using these preconditioners,
several additional issues are observed and alleviated to a certain
extent by applying the reduced domain approximation. MCSA solutions
for the fuel assembly problem are then verified by comparing the
solutions against production Krylov solvers for problems with varying
energy groups. Finally performance is analyzed through comparison with
those same Krylov solvers in terms of both iterative performance and
CPU timing using the same set of problems utilized for the
verification.

@@

    u_4 = 7\phi_6 \:.

@@

@@

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:multigroup_spn_equations}) with the fission
source derived in \S~\ref{sec:pn_eigenvalue}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with the fission matrix $\mathbf{F}$ given by
Eq~(\ref{eq:fission_matrix}). We again apply the change of variables
to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_n\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
It should be noted here that with respect to a general MCSA scheme,
the introduction of the fission in the system will only affect the
source vector in the linear system. The linear operator, and therefore
overall MCSA performance will be dictated by streaming and scattering
as defined on the left-hand side.

@@

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin materials and the associated grid is
given by Figure~\ref{fig:problem3_radial_mat} while a cross section of
the materials configuration along the horizontal axis is given by
Figure~\ref{fig:problem3_axial_mat}. A detailed view of the assembly
bottom is given in Figure~\ref{fig:problem3_end}. On the top and
bottom of the assembly, vacuum conditions are used as well as on the
top and right boundaries in
Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are used
on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized moderator. Each fuel pin is resolved by a $2 \times
      2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
Based on the success of block Jacobi preconditioning with the test
problem used for the spectral radius parameter study, we use it first
to solve the $17 \times 17$ fuel assembly problem. A single energy
group problem was first solved with $SP_1$ discretization, effectively
giving the one-speed neutron diffusion system for the fuel assembly
resulting in 20,088 degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of
MCSA. To investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor converge is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

@@

\subsection{MCSA Breakdown}
\label{subsec:mcsa_break_down}
For the fuel assembly problem, the initial block Jacobi
preconditioned\footnote{For 1-group $SP_1$ problems the block size is
  1, giving a point Jacobi preconditioner.}  calculations used a
one-speed $SP_1$ discretization, effectively giving a diffusion
system. To study the breakdown of MCSA at iteration matrix spectral
radii near one, we will use the simpler homogeneous 2-dimensional
one-speed neutron diffusion system presented in
Appendix~\ref{chap:diffusion_problem} to isolate this behavior. In
this system, we can vary the cross sections while maintaining a fixed
grid in order to achieve varying spectral radii. For these studies, we
neglect fission as MCSA behavior is dictated by the transport operator
$\mathbf{M}$ in an eigenvalue scheme with the fission matrix used to
generate a fixed source.

For each solver and estimator combination, the spectral radius of the
iteration matrix generated by the diffusion problem was varied by
changing the absorption cross section from 0.25 to 100 while fixing
the grid size at $100 \times 100$ with $h = 0.1$ and a fixed
scattering cross section of unity. For each parameter variation, a
minimum of one stochastic history per degree of freedom (DOF) in the
problem was used to compute the Monte Carlo correction. If the solver
could not converge in less than 100 iterations, the number of
histories was increased by increments of 5,000 until convergence was
achieved in less than 100 iterations. The number of iterations
required to converge MCSA and the time to converge was recorded as a
means to capture the breakdown.

@@

In addition to the significantly larger number of histories required
to achieve convergence for ill-conditioned problems another penalty is
paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in seconds
required to compute a single random walk averaged over the entire set
of histories run in the calculation over all iterations. As the
spectral radius increases (correlating to a higher ratio of scattering
in the system) the random walk lengths increase, using more CPU time
to finish the computation. Compared to spectral radii of 0.5, larger
spectral radii over 0.97 have histories that require two orders of
magnitude more computation time. This significant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is entirely ineffective and not practical. We
therefore require a more expansive set of preconditioning techniques
to move the eigenvalue spectrum of the $SP_N$ problem into a regime in
which MCSA is more applicable and in which performance is improved.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

@@

MCSA preconditioned with SPAINV-ILUT was also used to solve the same
fuel assembly problem as with the other preconditioners for the same
1-group $SP_1$ discretization. To study convergence sensitivity to
SPAINV-ILUT parameters, the ILUT drop tolerance and ILUT fill level
were parametrically varied with the number of iterations to converge a
single eigenvalue iteration for the fuel assembly problem recorded
along with the maximum number of non-zero entries observed in all
matrix rows for the left/right preconditioned composite linear
operator. SPAINV parameters for the data presented here were fixed as
it was found that the preconditioning quality was largely insensitive
to their variation. In addition, the fact that SPAINV is being used to
preconditioned a problem already preconditioned with ILUT means that
the input matrix sparsity pattern is already somewhat dense and large
SPAINV levels would result in an even denser system. Therefore, a
smaller SPAINV sparsity pattern level will be used for this
preconditioning. For each calculation, the number of iterations
required to converge reported was for a single eigenvalue iteration.
Again, \sn{3}{4} histories are used at each MCSA iteration to compute
the Monte Carlo correction using the adjoint collision estimator. All
SPAINV-ILUT calculations reported here were performed on a single CPU.
